---
title: "MSstats: Metabolomics workflow with MZMine"
date: June 17th, 2026
---


```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=7, warning=FALSE, message=FALSE)
options(width=110)
```

```{=html}
<!--
%\VignetteIndexEntry{MSstats: Metabolomics workflow with MZMine}
%\VignetteEngine{knitr::knitr}
-->
```
# __MSstats: Metabolomics workflow with MZMine__

Author: MSstats Team

Date: June 17th, 2026

## __Introduction__

`MSstats` supports differential analysis of metabolomics data acquired with
LC-MS untargeted workflows. This vignette walks an end-to-end run: import
MZMine feature quantifications and library annotations, layer in SIRIUS
structure identifications, convert to the MSstats format, summarize features
into compound-level abundance, and test for differences between conditions.

Compound identification combines two evidence sources:

* __MZMine compound names__ come from MS/MS spectral-library matching and
  correspond to MSI Level 2 putative identifications (Sumner et al., 2007).
* __SIRIUS names__ come from in-silico structure prediction and correspond to
  MSI Level 3 identifications. The SIRIUS pass extends discovery coverage to
  features the spectral library does not cover.

`MZMinetoMSstatsFormat` is re-exported from `MSstatsConvert`, so attaching
`MSstats` alone is enough to run the full workflow.

## __1. Setup__

```{r setup}
library(MSstats)
library(data.table)
```

## __2. Load example data__

Example MZMine input, sample annotation, MZMine library annotations, and
SIRIUS structure identifications ship with `MSstatsConvert` and are loaded
via `system.file()`.

```{r load-data}
input_path = system.file("tinytest/raw_data/MZMine/mzmine_input.csv",
                         package = "MSstatsConvert")
annotation_path = system.file("tinytest/raw_data/MZMine/annotation.csv",
                              package = "MSstatsConvert")
mzmine_ann_path = system.file("tinytest/raw_data/MZMine/mzmine_annotations.csv",
                              package = "MSstatsConvert")
sirius_path = system.file("tinytest/raw_data/MZMine/structure_identifications.tsv",
                          package = "MSstatsConvert")

mzmine_input = data.table::fread(input_path)
annotation = data.table::fread(annotation_path)
mzmine_annotations = data.table::fread(mzmine_ann_path)
sirius_annotations = data.table::fread(sirius_path)

head(mzmine_input, 5)
head(annotation)
head(mzmine_annotations)
head(sirius_annotations)
```

The MZMine feature table is wide: one row per feature, columns `row ID`,
`row m/z`, `row retention time`, and per-sample `"<run> Peak area"` columns.
The annotation table maps each MS run to its `Condition` and `BioReplicate`.
`mzmine_annotations` is the spectral-library match table
(`id`, `compound_name`, `score`, `adduct`); features with multiple library
hits resolve to the highest-scoring compound. `sirius_annotations` is
SIRIUS's `structure_identifications.tsv`; its `mappingFeatureId` joins to
`row ID` in the MZMine input.

## __3. Convert with `MZMinetoMSstatsFormat`__

```{r convert, message = FALSE}
mzmine_msstats = MZMinetoMSstatsFormat(
    input              = mzmine_input,
    annotation         = annotation,
    mzmine_annotations = mzmine_annotations,
    sirius_annotations = sirius_annotations,
    use_log_file       = FALSE
)
head(mzmine_msstats)
```

`ProteinName` is assigned per feature in priority order: (1) the
highest-scoring MZMine compound name when present, (2) the SIRIUS name when
MZMine has no match, (3) an `m/z_RT` fallback identifier for features
neither source identified. Every feature is retained -- discovery coverage
is preserved at the cost of a wider multiple-testing burden in Section 5.

### Lactate caveat

Lactate (feature 3) is missing one of its four measurements in this fixture, so its
differential result is unreliable. That value is dropped rather than estimated, so Lactate is
tested on three points and its degrees of freedom fall to 1, against 2 for the fully measured
compounds. With so little data the variance estimate is unstable, which is why Lactate shows a
very small standard error, a large t-statistic, and the only small p-value in the table. Treat
it as an artifact of the tiny example, not a real difference.

## __4. Summarize with `dataProcess`__

```{r summarize, message = FALSE}
summarized = dataProcess(
    mzmine_msstats,
    logTrans      = 2,
    normalization = "equalizeMedians",
    featureSubset = "all",
    summaryMethod = "TMP",
    censoredInt   = "NA",
    MBimpute      = TRUE,
    use_log_file  = FALSE
)
head(summarized$FeatureLevelData)
head(summarized$ProteinLevelData)
```

The settings above mirror the `MSstatsWorkflow` vignette: log-2 transform,
median-equalized normalization, all features used, and Tukey median polish
summarization. Model-based imputation is enabled (`MBimpute = TRUE`), but no
values are imputed in this small example. Caffeine is detected at two adducts (`[M+H]+` on feature 1,
`[M+Na]+` on feature 6) and is summarized into a single compound-level
abundance per run.

## __5. Test for differences with `groupComparison`__

With two conditions in the design, a single Control-vs-Treatment contrast
is generated by passing `"pairwise"`:

```{r contrast, message = FALSE}
comparison = groupComparison(contrast.matrix = "pairwise",
                             data            = summarized,
                             use_log_file    = FALSE)
comparison$ComparisonResult
```

Each row of `ComparisonResult` is one compound (or `m/z_RT` fallback) tested
against the contrast. Columns of interest: `log2FC`, `pvalue`, and
`adj.pvalue`. The `issue` column flags compounds that could not be tested
normally, for example one missing from an entire condition; in this small
fixture it is empty for every compound shown.

## __6. Visualization__

Profile plots show feature-level intensities alongside the protein-level
summary. Caffeine is identified at two adducts in this dataset and is
summarized into a single compound -- the profile plot makes that aggregation
visible.

```{r profile, fig.width = 8, fig.height = 5}
dataProcessPlots(summarized,
                 type          = "ProfilePlot",
                 which.Protein = "Caffeine",
                 address       = FALSE)
```

For a study-wide view of fold-change versus significance, pass the
`groupComparison` result to `groupComparisonPlots`. On a four-sample fixture
the volcano is sparse; on a real metabolomics dataset it is the standard
summary plot:

```{r volcano, eval = FALSE}
groupComparisonPlots(data    = comparison$ComparisonResult,
                     type    = "VolcanoPlot",
                     address = FALSE)
```

## __References__

Sumner LW, Amberg A, Barrett D, et al. (2007). Proposed minimum reporting standards for
chemical analysis: Chemical Analysis Working Group (CAWG) Metabolomics Standards Initiative
(MSI). *Metabolomics* 3(3): 211-221. doi: [10.1007/s11306-007-0082-2](https://doi.org/10.1007/s11306-007-0082-2)

```{r session}
sessionInfo()
```